---
title: "Getting and Cleaning Data - Course Project"
date: "November 22, 2015"
output: html_document
---

# Getting and Cleaning Data - Course Project

### Introduction

This document covers processing of data of human activity recognition based on smartfone collected data. Data is avialable under URL:
https://d396qusza40orc.cloudfront.net/getdata%2Fprojectfiles%2FUCI%20HAR%20Dataset.zip  The full descriptiois available under URL:   http://archive.ics.uci.edu/ml/datasets/Human+Activity+Recognition+Using+Smartphones.

The goal of the processing was to create a tidy dataset containing accordin to following points
- Merge the training and the test sets to create one data set.  
- Extract only the measurements on the mean and standard deviation for each measurement  
- Use descriptive activity names to name the activities in the data set  
- Appropriately label the data set with descriptive variable names.  
- Create a second, independent tidy data set with the average of each variable for each activity and each subject.  

**Data Set Information**

The experiments have been carried out with a group of 30 volunteers within an age bracket of 19-48 years. Each person performed six activities (WALKING, WALKING_UPSTAIRS, WALKING_DOWNSTAIRS, SITTING, STANDING, LAYING) wearing a smartphone (Samsung Galaxy S II) on the waist. Using its embedded accelerometer and gyroscope, we captured 3-axial linear acceleration and 3-axial angular velocity at a constant rate of 50Hz. The experiments have been video-recorded to label the data manually. The obtained dataset has been randomly partitioned into two sets, where 70% of the volunteers was selected for generating the training data and 30% the test data. 

The sensor signals (accelerometer and gyroscope) were pre-processed by applying noise filters and then sampled in fixed-width sliding windows of 2.56 sec and 50% overlap (128 readings/window). The sensor acceleration signal, which has gravitational and body motion components, was separated using a Butterworth low-pass filter into body acceleration and gravity. The gravitational force is assumed to have only low frequency components, therefore a filter with 0.3 Hz cutoff frequency was used. From each window, a vector of features was obtained by calculating variables from the time and frequency domain.

**Attribute Information**

For each record in the dataset it is provided
- Triaxial acceleration from the accelerometer (total acceleration) and the estimated body acceleration.     
- Triaxial Angular velocity from the gyroscope.   
- A 561-feature vector with time and frequency domain variables.   
- Its activity label.   
- An identifier of the subject who carried out the experiment.  

### Files used in the process

In this activity we will use following data files
- features.txt - FeaturesNames - file with labels of features (measurments)    
- Y_test.txt - ActivityTest - testing set containing codes of activities  
- Y_train.txt - ActivityTrain - training set containing codes of activities  
- subject_test.txt - SubjectTest - testing set containing codes of users   
- subject_train.txt - SubjectTrain - training set containing codes of users  
- X_test.txt - FeaturesTest - testing set containing measurements   
- X_train.txt - FeaturesTrain - training set containing measurements   
- activity_labels.txt - ActivityLabels - file with labels of activity - will be used to "decode" activity numbers

### 1. Merge the training and the test sets to create one data set.

First we need to download and unpack the data

```{r}
if(!file.exists("./data")){dir.create("./data")}
fileUrl <- "https://d396qusza40orc.cloudfront.net/getdata%2Fprojectfiles%2FUCI%20HAR%20Dataset.zip"
download.file(fileUrl,destfile="./data/data.zip")
unzip(zipfile="./data/data.zip",exdir="./data")
```

Then we need to read nacessary data and combine into one dataset.

```{r}
FeaturesNames <- read.table("./data/UCI HAR Dataset/features.txt",head=FALSE)

ActivityTest  <- read.table("./data/UCI HAR Dataset/test/Y_test.txt",header = FALSE)
ActivityTrain  <- read.table("./data/UCI HAR Dataset/train/Y_train.txt",header = FALSE)

SubjectTest  <- read.table("./data/UCI HAR Dataset/test/subject_test.txt",header = FALSE)
SubjectTrain  <- read.table("./data/UCI HAR Dataset/train/subject_train.txt",header = FALSE)

FeaturesTest  <- read.table("./data/UCI HAR Dataset/test/X_test.txt",header = FALSE)
FeaturesTrain  <- read.table("./data/UCI HAR Dataset/train/X_train.txt",header = FALSE)

Activity <- rbind(ActivityTest,ActivityTrain)
Subject <- rbind(SubjectTest,SubjectTrain)
Features <- rbind(FeaturesTest,FeaturesTrain)

ActivityTest <- NULL
ActivityTrain <- NULL
SubjectTest  <- NULL
SubjectTrain <- NULL
FeaturesTest  <- NULL
FeaturesTrain <- NULL

names(Activity) <- c("Activity")
names(Subject) <- c("Subject")
names(Features) <- FeaturesNames[,2]

CombinedData <- cbind(Subject,Activity,Features)

Activity <- NULL
Subject <- NULL
Features  <- NULL
FeaturesNames  <- NULL
```

### 2. Extract only the measurements on the mean and standard deviation for each measurement.

Then we need to select only mean and standard deviation columns. 

```{r}
MeanSTDcolnames <- colnames(CombinedData[grepl("mean|std",colnames(CombinedData))])
CombinedData <- CombinedData[,c("Subject","Activity",MeanSTDcolnames)]
```

### 3. Use descriptive activity names to name the activities in the data set

```{r}
ActivityLabels <- read.table("./data/UCI HAR Dataset/activity_labels.txt",header = FALSE)

CombinedData <- merge(CombinedData,ActivityLabels,by.x = "Activity",by.y = "V1")
CombinedData$Activity <- CombinedData$V2
CombinedData$V2 <- NULL
ActivityLabels  <- NULL
```

### 4. Appropriately label the data set with descriptive variable names.

```{r}
NewColNames <- names(CombinedData)
NewColNames <- gsub('\\(|\\)',"",NewColNames)
NewColNames <- gsub("Acc", "Acceleration ", NewColNames)
NewColNames <- gsub("Mag", "Magnitude ", NewColNames)
NewColNames <- gsub("Gyro", "Gyroscope ", NewColNames)
NewColNames <- gsub("Gravity", "Gravity ", NewColNames)
NewColNames <- gsub("^t", "Time ", NewColNames)
NewColNames <- gsub("^f", "Frequency ", NewColNames)
NewColNames <- gsub("BodyBody", "Body", NewColNames)
NewColNames <- gsub("Body", "Body ", NewColNames)
NewColNames <- gsub("Jerk", "Jerk ", NewColNames)
NewColNames <- gsub("-mean", "- Mean ", NewColNames)
NewColNames <- gsub("-std", "- STD.VAR ", NewColNames)
names(CombinedData) <- NewColNames
```

### 5. Create a second, independent tidy data set with the average of each variable for each activity and each subject.

```{r}
TidyAvgData <- aggregate(.~Subject+Activity,CombinedData,mean)
write.table(TidyAvgData, file = "tidy.txt", row.names = FALSE)
```

